## PDAC combined locus level index + EM + GeneFull + aggregated result

```{r}

#BiocManager::install("SingleCellSignalR")
#library(SingleR)
#remotes::install_github("cvarrichio/Matrix.utils")
#install.packages("~/R/library-ood-4.0.4/Matrix_1.6-5.tgz", repos = NULL, type = .Platform$pkgType)
#install.packages("~/R/Matrix_1.6-4.tar.gz", repos = NULL, type="source")
library("Matrix")
#if (!require("BiocManager", quietly = TRUE)) 
#  install.packages("BiocManager") 
#BiocManager::install(version = "3.14")
library("BiocManager")
#BiocManager::install("SeuratObject")
library(ggrepel)
library(dplyr)
library(tidyverse)
library(patchwork)
library(ggplot2)
#library(DropletUtils)
library("stringr")
library("harmony")
library(RColorBrewer)
library(Matrix)
library(SeuratObject)
library(Seurat)
library(data.table)
#BiocManager::install("DropletUtils")
library("SeuratObject")
#BiocManager::install("Seurat") 
library("Seurat")


#BiocManager::install("DropletUtils")
```

##after running Step0_filtering_integration_repCDsep.Rmd
```{r}
load("rdata/pdac/PDAC_EMGeneFull_filtNorm.RData")

#data.filt_normed = FindVariableFeatures(data.filt_normed, verbose = F)
## scale the data
combined = ScaleData(combined, vars.to.regress = c("nFeature_RNA", "percent.mt"), verbose = F)
## PCA
combined = RunPCA(combined, verbose = F, npcs = 20)
combined = RunUMAP(combined, dims = 1:10, verbose = F)
# Examine and visualize PCA results a few different ways
print(combined[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(combined, dims = 1:2, reduction = "pca")

```

##add sample info to metadata
```{r}
merged_seurat = combined
# Extract the SRR part from orig.ident

cond_map <- c(
"SRR30218424"= "Pt13_female_Peritoneal metastasis_FOLFIRINOX\\, Capecitabine\\, gemcitabine/abraxane/ denosumab",
"SRR30218425"= "Pt13_female_Peritoneal metastasis_FOLFIRINOX\\, Capecitabine\\, gemcitabine/abraxane/ denosumab",
"SRR30218426"= "Pt13_female_Lung metastasis_FOLFIRINOX\\, Capecitabine\\, gemcitabine/abraxane/ denosumab",
"SRR30218427"= "Pt13_female_Liver metastasis_FOLFIRINOX\\, Capecitabine\\, gemcitabine/abraxane/ denosumab",
"SRR30218428"= "Pt13_female_Primary PDAC_FOLFIRINOX\\, Capecitabine\\, gemcitabine/abraxane/ denosumab",
"SRR30218429"= "Pt12_female_Peritoneal metastasis_gemcitabine",
"SRR30218430"= "Pt12_female_Lung metastasis_gemcitabine",
"SRR30218431"= "Pt12_female_Liver metastasis_gemcitabine",
"SRR30218432"= "Pt12_female_Primary PDAC_gemcitabine",
"SRR30218433"= "Pt11_male_Peritoneal metastasis_gemcitabine\\, Oxaliplatin\\, leucovorin-modulated 5-FU\\, irinotecan\\",
"SRR30218434"= "Pt11_male_Lung metastasis_gemcitabine\\, Oxaliplatin\\, leucovorin-modulated 5-FU\\, irinotecan\\",
"SRR30218435"= "Pt11_male_Liver metastasis_gemcitabine\\, Oxaliplatin\\, leucovorin-modulated 5-FU\\, irinotecan\\",
"SRR30218436"= "Pt11_male_Primary PDAC_gemcitabine\\, Oxaliplatin\\, leucovorin-modulated 5-FU\\, irinotecan\\",
"SRR30218437"= "Pt10_male_Peritoneal metastasis_FOLFIRINOX\\, carboplatin/FOLFOX\\, gemcitabine/abraxane",
"SRR30218438"= "Pt10_male_Lung metastasis_FOLFIRINOX\\, carboplatin/FOLFOX\\, gemcitabine/abraxane",
"SRR30218439"= "Pt10_male_Liver metastasis_FOLFIRINOX\\, carboplatin/FOLFOX\\, gemcitabine/abraxane",
"SRR30218440"= "Pt10_male_Primary PDAC_FOLFIRINOX\\, carboplatin/FOLFOX\\, gemcitabine/abraxane",
"SRR30218441"= "Pt9_female_Peritoneal metastasis_Cisplatin\\, gemcitabine\\, ADH-1\\, 5-FU (Clinical trial)",
"SRR30218442"= "Pt9_female_Liver metastasis_Cisplatin\\, gemcitabine\\, ADH-1\\, 5-FU (Clinical trial)",
"SRR30218443"= "Pt9_female_Liver metastasis_Cisplatin\\, gemcitabine\\, ADH-1\\, 5-FU (Clinical trial)",
"SRR30218444"= "Pt9_female_Primary PDAC_Cisplatin\\, gemcitabine\\, ADH-1\\, 5-FU (Clinical trial)",
"SRR30218445"= "Pt8_male_Peritoneal metastasis_FOLFIRINOX\\, FOLFIRI\\, Abraxane/Gemcitabine",
"SRR30218446"= "Pt8_male_Liver metastasis_FOLFIRINOX\\, FOLFIRI\\, Abraxane/Gemcitabine",
"SRR30218447"= "Pt8_male_Liver metastasis_FOLFIRINOX\\, FOLFIRI\\, Abraxane/Gemcitabine",
"SRR30218448"= "Pt8_male_Primary PDAC_FOLFIRINOX\\, FOLFIRI\\, Abraxane/Gemcitabine",
"SRR30218449"= "Pt7_male_Peritoneal metastasis_gemcitabine\\, ADH1\\, cisplatin",
"SRR30218450"= "Pt7_male_Liver metastasis_gemcitabine\\, ADH1\\, cisplatin",
"SRR30218451"= "Pt7_male_Liver metastasis_gemcitabine\\, ADH1\\, cisplatin",
"SRR30218452"= "Pt7_male_Primary PDAC_gemcitabine\\, ADH1\\, cisplatin",
"SRR30218453"= "Pt6_male_Peritoneal metastasis_gemcitabine\\, AVN944\\, erlotinib",
"SRR30218454"= "Pt6_male_Peritoneal metastasis_gemcitabine\\, AVN944\\, erlotinib",
"SRR30218455"= "Pt6_male_Liver metastasis_gemcitabine\\, AVN944\\, erlotinib",
"SRR30218456"= "Pt6_male_Primary PDAC_gemcitabine\\, AVN944\\, erlotinib",
"SRR30218457"= "Pt5_female_Liver metastasis_No treatment",
"SRR30218458"= "Pt5_female_Liver metastasis_No treatment",
"SRR30218459"= "Pt5_female_Primary PDAC_No treatment",
"SRR30218460"= "Pt4_female_Peritoneal metastasis_No treatment",
"SRR30218461"= "Pt4_female_Lung metastasis_No treatment",
"SRR30218462"= "Pt4_female_Liver metastasis_No treatment",
"SRR30218463"= "Pt4_female_Primary PDAC_No treatment",
"SRR30218464"= "Pt3_female_Peritoneal metastasis_No treatment",
"SRR30218465"= "Pt3_female_Liver metastasis_No treatment",
"SRR30218466"= "Pt3_female_Primary PDAC_No treatment",
"SRR30218467"= "Pt2_female_Peritoneal metastasis_No treatment",
"SRR30218468"= "Pt2_female_Lung metastasis_No treatment",
"SRR30218469"= "Pt2_female_Primary PDAC_No treatment",
"SRR30218470"= "Pt1_female_Peritoneal metastasis_No treatment",
"SRR30218471"= "Pt1_female_Liver metastasis_No treatment",
"SRR30218472"= "Pt1_female_Primary PDAC_No treatment"
)

merged_seurat$sample = merged_seurat$orig.ident

merged_seurat$orig.ident <- unname(cond_map[merged_seurat$sample])

merged_seurat@meta.data <- merged_seurat@meta.data %>%
  separate(orig.ident, into = c("patient", "gender", "tissue", "treatment"), sep = "_", remove = FALSE)


# Create metadata dataframe
metadata <- merged_seurat@meta.data
# Add cell IDs to metadata
metadata$cells <- rownames(metadata)
# Rename columns
metadata <- metadata %>%
        dplyr::rename(Cond = orig.ident)

# Add metadata back to Seurat object
merged_seurat@meta.data <- metadata
Idents(merged_seurat) <- "Cond"    
                           
VlnPlot(merged_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
plot1 <- FeatureScatter(merged_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(merged_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

```

```{r}
library(RColorBrewer)
library(ggplot2)
data.filt_normed = merged_seurat
my_colors1 <- colorRampPalette(brewer.pal(12, "Paired"))(50) ##if more than 24 colors needed
my_colors <- c(rev(brewer.pal(n = 12, name = "Paired")),rev(brewer.pal(n = 12, name = "Set2")))


DimPlot(data.filt_normed, reduction = "umap", group.by = "Cond", repel = FALSE, cols = my_colors1) +
  guides(color = guide_legend(ncol = 1)) +  # Legend in 1 column
  theme(legend.position = "right")        

#see different umaps
DimPlot(data.filt_normed, reduction = "umap", group.by = "patient", label = FALSE, repel = TRUE,cols =my_colors)
DimPlot(data.filt_normed, reduction = "umap", group.by = "treatment", label = FALSE, repel = TRUE,cols =my_colors)
DimPlot(data.filt_normed, reduction = "umap", group.by = "tissue", label = FALSE, repel = TRUE,cols =my_colors)
DimPlot(data.filt_normed, reduction = "umap", group.by = "gender", label = TRUE, repel = TRUE,cols =my_colors)

```

## batch correction
```{r}
## add annotations in the original data for later correction
## By cellline
seuratObj_cellline <- RunHarmony(data.coding_filt_normed, "patient")
#DimPlot(object = seuratObj, reduction = "harmony", pt.size = .1, group.by = "seq_folder",  label = TRUE, repel = TRUE )
seuratObj_cellline <- RunUMAP(seuratObj_cellline, reduction = "harmony", dims = 1:10, verbose = F)

#save(seuratObj_cellline, file=paste0("./PDAC_batchCorrected.RData"))

```
## plot of BCed data
```{r}
#load("./PDAC_batchCorrected.RData")

DimPlot(seuratObj_cellline, reduction = "harmony", group.by = c("Cond"), label = F, repel = TRUE, cols = my_colors1)
DimPlot(seuratObj_cellline, reduction = "umap", group.by = "Cond", label = F, repel = F,cols = my_colors1) +
  guides(color = guide_legend(ncol = 1)) +  # Legend in 1 column
  theme(legend.position = "right")        


DimPlot(seuratObj_cellline, reduction = "umap", group.by = "patient", label = F, repel = F,cols = my_colors)
DimPlot(seuratObj_cellline, reduction = "umap", group.by = "tissue", label = F, repel = F,cols = my_colors)
DimPlot(seuratObj_cellline, reduction = "umap", group.by = "treatment", label = F, repel = F,cols = my_colors)
DimPlot(seuratObj_cellline, reduction = "umap", group.by = "gender", label = F, repel = F,cols = my_colors)


```
## clustering
```{r}
# check graph objects
seuratObj_cellline@graphs
#seuratObj_cellline <- FindNeighbors(seuratObj_cellline, dims = 1:6, reduction = "pca")
seuratObj_cellline <- FindNeighbors(seuratObj_cellline, dims = 1:10, reduction = "harmony")
# Find clusters
seuratObj_cellline <- FindClusters(seuratObj_cellline, resolution = 0.5) ## 0.1 very coarse, 0.5 moderate, 0.8 Fine
# View cluster assignments
head(Idents(seuratObj_cellline))
# Visualize clusters
my_colors <- c(brewer.pal(n=12, name = "Set3"),brewer.pal(n=2, name = "Set1"))
#DimPlot(seuratObj_cellline, reduction = "pca", group.by = "seq_folder", label = F,cols = my_colors)
DimPlot(seuratObj_cellline, reduction = "umap", group.by = "seurat_clusters", label = TRUE,cols = my_colors)


```

## cluster markers
```{r}
seuratObj_cellline <- JoinLayers(seuratObj_cellline) 
# Identify marker genes for each cluster
cluster_markers <- FindAllMarkers(seuratObj_cellline, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
# View top markers for each cluster
head(cluster_markers)
cluster_markers$gene <- sub(".*_", "", cluster_markers$gene)
```
##look at cluster markers
```{r}
#top 10 most expressed per cluster
top10 <- cluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC)

seuratObj_cellline <- ScaleData(seuratObj_cellline, features = unique(top10$gene))

DoHeatmap(seuratObj_cellline,
          features = unique(top10$gene),
          group.by = "seurat_clusters",
          cells = sample(colnames(seuratObj_cellline), 1000)) +
  theme(axis.text.y = element_text(size = 10))

```

### cell annotation per cluster ---- using GPT4
```{R}
# IMPORTANT! Assign your OpenAI API key. See Vignette for details
#put in your personal gpt API
#Sys.setenv(OPENAI_API_KEY ='...')

             
#install.packages("devtools")  # if not already installed
#install.packages("openai")
#remotes::install_github("Winnie09/GPTCelltype")

# Load packages
library(GPTCelltype) 
library(openai)
# Assume you have already run the Seurat pipeline https://satijalab.org/seurat/
# "obj" is the Seurat object; "markers" is the output from FindAllMarkers(obj)
# Cell type annotation by GPT-4 
res_10 <- gptcelltype(cluster_markers, model = 'gpt-4',topgenenumber = 10)
res_20 <- gptcelltype(cluster_markers, model = 'gpt-4',topgenenumber = 20)
res_30 <- gptcelltype(cluster_markers, model = 'gpt-4',topgenenumber = 30)
res_40 <- gptcelltype(cluster_markers, model = 'gpt-4',topgenenumber = 40)
# Assign cell type annotation back to Seurat object
seuratObj_cellline@meta.data$gptcelltype_10<- as.factor(res_10[as.character(Idents(seuratObj_cellline))])
seuratObj_cellline@meta.data$gptcelltype_20 <- as.factor(res_20[as.character(Idents(seuratObj_cellline))])
seuratObj_cellline@meta.data$gptcelltype_30 <- as.factor(res_30[as.character(Idents(seuratObj_cellline))])
seuratObj_cellline@meta.data$gptcelltype_40 <- as.factor(res_40[as.character(Idents(seuratObj_cellline))])
# Visualize cell type annotation on UMAP
my_colors <- c(rev(brewer.pal(n = 12, name = "Paired")),rev(brewer.pal(n = 12, name = "Set2")))
DimPlot(seuratObj_cellline,group.by='gptcelltype_10', cols=my_colors)
DimPlot(seuratObj_cellline,group.by='gptcelltype_20', cols=my_colors)
DimPlot(seuratObj_cellline,group.by='gptcelltype_30', cols=my_colors)
DimPlot(seuratObj_cellline,group.by='gptcelltype_40', cols=my_colors)
seuratObj_cellline_anno = seuratObj_cellline
saveRDS(seuratObj_cellline, file = "./PDAC_cell_annot.rds")
saveRDS(seuratObj_cellline, file = "./0724PDAC_EM_GENEFULL_cellAno.rds")


```
### SingleR
```{r}
############################ Immune reference

ref.data <- celldex::ImmGenData()

# Run SingleR
pred.sc <- SingleR(
  test = GetAssayData(seuratObj_cellline, slot = "data"),
  ref = ref.data,
  labels = ref.data$label.main # Ensure the correct label column
)

seuratObj_cellline$SingleR_ImmuneGen <- pred.sc$labels
```
### celltype annotation per cluster ---- PANC data
```{r}
library(scRNAseq)
############################ PDAC reference
ref <- BaronPancreasData(which = "human")
assayNames(ref)
assay(ref, "logcounts") <- log1p(counts(ref))

# Run SingleR
pred.sc <- SingleR(
  test = GetAssayData(seuratObj_cellline, slot = "data"),
  ref = ref,
  labels = ref$label # Ensure the correct label column #label.main
)

seuratObj_cellline$SingleR_Panc <- pred.sc$labels

```
### celltype annotation per cluster ---- Mouse RNA ref
```{r}

############################# Mouse RNA reference
ref <- celldex::HumanPrimaryCellAtlasData()

singleR_results <- SingleR(test = GetAssayData(seuratObj_cellline, slot = "data"),
                           ref = ref,
                           labels = ref$label.main)

# Add annotations to Seurat metadata
seuratObj_cellline$SingleR_mouseRNA <- singleR_results$labels
```
#save annotation and visualize
```{r}
saveRDS(seuratObj_cellline, file = "./seurat_annot.rds")
```
```{r}
DimPlot(seuratObj_cellline,group.by='SingleR_ImmuneGen', cols=my_colors1)
DimPlot(seuratObj_cellline,group.by='SingleR_mouseRNA', cols=my_colors1)

DimPlot(seuratObj_cellline,group.by='gptcelltype_10', cols=my_colors)
DimPlot(seuratObj_cellline,group.by='seurat_clusters', cols=my_colors,)

seuratObj_cellline$seurat_clusters <- as.character(seuratObj_cellline$seurat_clusters)
```

##correlation matrix
```{r}
library(ggplot2)
library(reshape2)

seurat_obj = readRDS("seurat_annot.rds")
repeat_class = fread("/data1/greenbab/users/nij2/ref/all_landscape_rm.csv")
##optional, manually combining two same cell type annotations
seurat_obj$gptcelltype_10 = trimws(seurat_obj$gptcelltype_10)
seurat_obj$gptcelltype_10[seurat_obj$gptcelltype_10 == "Liver Cells "] <- "Liver Cells"
DimPlot(seurat_obj,group.by='gptcelltype_10', cols=my_colors)

#find repeats
repeat_genes <- intersect(rownames(seurat_obj), repeat_class$gene.id)
repeat_table_filtered <- repeat_class[repeat_class$gene.id %in% repeat_genes, ]
repeat_class_list <- split(repeat_table_filtered$gene.id, repeat_table_filtered$class)

# Add module scores for each repeat class
seurat_obj <- AddModuleScore(seurat_obj, features = repeat_class_list, name = "RepeatClass")
#rename colnames
repeat_class_names <- names(repeat_class_list)
score_cols <- grep("^RepeatClass[0-9]+$", colnames(seurat_obj@meta.data), value = TRUE)
colnames(seurat_obj@meta.data)[match(score_cols, colnames(seurat_obj@meta.data))] <- repeat_class_names

#repeat matrix
# Extract repeat module score matrix: rows = repeats, columns = cells
repeat_mat <- t(seurat_obj@meta.data[, repeat_class_names]) 
```
#tissue binary matrix vs. repeat module score matrix
```{r}
#tissue type
meta <- seurat_obj@meta.data
meta$cells <- colnames(seurat_obj)
rownames(meta) <- meta$cells        # set rownames to cell names
#create tissue binary matrix
tissue_mat <- model.matrix(~ 0 + tissue, data = meta)
colnames(tissue_mat) <- gsub("tissue", "", colnames(tissue_mat))  #  cleanup
tissue_mat <- t(tissue_mat)  # now rows = tissue types, cols = cells
#graph
cor_mat <- cor(t(repeat_mat), t(tissue_mat), use = "pairwise.complete.obs")
pheatmap(cor_mat, cluster_rows = TRUE, cluster_cols = TRUE)

#dotplot
cor_mat = cor_mat[c("DNA", "LINE", "SINE", "LTR", "rRNA", "Satellite", "Simple_repeat", "sRNA", "tRNA"), ]
cor_df <- melt(cor_mat)
colnames(cor_df) <- c("RepeatClass", "Type", "Correlation")

ggplot(cor_df, aes(x = Type, y = RepeatClass, color = Correlation, size = abs(Correlation))) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_classic() +
  labs(title = "Correlation between Repeat Classes and Cell Types",
       x = "Cell Type", y = "Repeat Class", color = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#vln plot 
score_df <- seurat_obj@meta.data[, c("tissue", repeat_class_names)]
score_long <- score_df %>%
  pivot_longer(cols = -tissue, names_to = "RepeatClass", values_to = "ModuleScore")

selected_classes <- c("DNA", "LINE", "SINE", "LTR", "rRNA", "Satellite", "Simple_repeat", "sRNA", "tRNA")

score_long_filtered <- score_long %>% 
  filter(RepeatClass %in% selected_classes)

ggplot(score_long_filtered, aes(x = tissue, y = ModuleScore, fill = tissue)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~ RepeatClass, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = "Tissue", y = "Module Score", title = "Repeat Class Expression by Tissue")

ggplot(score_long_filtered, aes(x = RepeatClass, y = ModuleScore, fill = RepeatClass)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~ tissue, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Repeat Class", y = "Module Score", title = "Repeat Class Expression by Tissue")


```

#treatment binary matrix vs. repeat module score
```{r}
#create treatment binary matrix
treatment_mat <- model.matrix(~ 0 + treatment, data = meta)
colnames(treatment_mat) <- gsub("treatment", "", colnames(treatment_mat))  #  cleanup
treatment_mat <- t(treatment_mat)  # now rows = tissue types, cols = cells
#graph
cor_mat <- cor(t(repeat_mat), t(treatment_mat), use = "pairwise.complete.obs")
pheatmap(cor_mat, cluster_rows = TRUE, cluster_cols = TRUE)

#dotplot
cor_mat = cor_mat[c("DNA", "LINE", "SINE", "LTR", "rRNA", "Satellite", "Simple_repeat", "sRNA", "tRNA"), ]
cor_df <- melt(cor_mat)
colnames(cor_df) <- c("RepeatClass", "Type", "Correlation")

ggplot(cor_df, aes(x = Type, y = RepeatClass, color = Correlation, size = abs(Correlation))) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_classic() +
  labs(title = "Correlation between Repeat Classes and Cell Types",
       x = "Cell Type", y = "Repeat Class", color = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#cell type binary matrix
```{R}
#cell type
meta$celltype <- as.character(meta$gptcelltype_10)
meta <- meta[!is.na(meta$gptcelltype_10), ]
celltype_mat <- model.matrix(~ 0 + celltype, data = meta)
colnames(celltype_mat) <- gsub("celltype", "", colnames(celltype_mat))  # optional cleanup
celltype_mat <- t(celltype_mat)  # rows = cell types, columns = cells
#NAs in cell type, only find common
common_cells <- intersect(colnames(repeat_mat), colnames(celltype_mat))
repeat_mat_aligned <- repeat_mat[, common_cells]
celltype_mat_aligned <- celltype_mat[, common_cells]
#graph
cor_mat <- cor(t(repeat_mat_aligned), t(celltype_mat_aligned), use = "pairwise.complete.obs")
pheatmap(cor_mat, cluster_rows = TRUE, cluster_cols = TRUE)


#dotplot
cor_mat = cor_mat[c("DNA", "LINE", "SINE", "LTR", "rRNA", "Satellite", "Simple_repeat", "sRNA", "tRNA"), ]
cor_df <- melt(cor_mat)
colnames(cor_df) <- c("RepeatClass", "Type", "Correlation")

ggplot(cor_df, aes(x = Type, y = RepeatClass, color = Correlation, size = abs(Correlation))) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_classic() +
  labs(title = "Correlation between Repeat Classes and Cell Types",
       x = "Cell Type", y = "Repeat Class", color = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

#mat for signature genes paper provided
```{R}
library(readxl)
cell_sig = read_excel("/data1/greenbab/users/nij2/ref/PDACtable9.xlsx")

sig_genes <- intersect(rownames(seurat_obj), cell_sig$gene.id)
sig_table_filtered <- cell_sig[cell_sig$gene.id %in% sig_genes, ]
sig_class_list <- split(sig_table_filtered$gene.id, sig_table_filtered$cell)

# Add module scores for each repeat class
seurat_obj <- AddModuleScore(seurat_obj, features = sig_class_list, name = "Signature")
#rename colnames
sig_class_names <- names(sig_class_list)
score_cols <- grep("^Signature[0-9]+$", colnames(seurat_obj@meta.data), value = TRUE)
colnames(seurat_obj@meta.data)[match(score_cols, colnames(seurat_obj@meta.data))] <- sig_class_names

# Extract repeat module score matrix: rows = repeats, columns = cells
signature_mat <- t(seurat_obj@meta.data[, sig_class_names]) 

cor_mat <- cor(t(repeat_mat), t(signature_mat), use = "pairwise.complete.obs")
cor_mat = cor_mat[c("DNA", "LINE", "SINE", "LTR"), ]
pheatmap(cor_mat, cluster_rows = TRUE, cluster_cols = TRUE)

```
#other pdac signature genes from https://www.nature.com/articles/s41467-023-36296-4
```{r}
markers = fread("/data1/greenbab/users/nij2/ref/cell_marker_pdac.csv")
seurat_obj = seuratObj_cellline

sig_genes <- intersect(rownames(seurat_obj), markers$gene.id)
sig_table_filtered <- markers[markers$gene.id %in% sig_genes, ]
sig_class_list <- split(sig_table_filtered$gene.id, sig_table_filtered$cell)

# Add module scores for each repeat class
seurat_obj <- AddModuleScore(seurat_obj, features = sig_class_list, name = "Signature")
#rename colnames
sig_class_names <- names(sig_class_list)
score_cols <- grep("^Signature[0-9]+$", colnames(seurat_obj@meta.data), value = TRUE)
colnames(seurat_obj@meta.data)[match(score_cols, colnames(seurat_obj@meta.data))] <- sig_class_names

# Extract repeat module score matrix: rows = repeats, columns = cells
signature_mat <- t(seurat_obj@meta.data[, sig_class_names]) 

cor_mat <- cor(t(repeat_mat), t(signature_mat), use = "pairwise.complete.obs")
cor_mat = cor_mat[c("DNA", "LINE", "SINE", "LTR"), ]
pheatmap(cor_mat, cluster_rows = TRUE, cluster_cols = TRUE)

saveRDS(seuratObj_cellline, file = "./0725PDAC_EM_GENEFULL_cellAno.rds")
```


##zoom into SINEs and LINEs
```{r}
library(dplyr)
library(stringr)


seurat_obj = readRDS("0725PDAC_EM_GENEFULL_cellAno.rds")
gene_names <- rownames(seurat_obj)

repeat_family <- sapply(gene_names, function(g) {
  if (grepl("_LINE$", g)) {
    return("x")  # Generic LINE family
  } else if (grepl("^Alu[YJSM]", g)) {
    return(sub("^([A-Za-z]+).*", "\\1", g))  # AluY, AluJ, etc.
  } else if (grepl("^L[1-5]", g)) {
    return(sub("^(L[1-5]).*", "\\1", g))  # L1PA2 → L1
  } else if (grepl("^CR1", g)) {
    return("CR1")
  } else if (grepl("^HAL1", g)) {
    return("HAL1")
  } else {
    return(NA)
  }
})

repeat_family_df <- data.frame(
  gene = gene_names,
  family = repeat_family,
  stringsAsFactors = FALSE
)
repeat_family_df$family <- case_when(
  str_detect(repeat_family_df$gene, "^AluY") ~ "AluY",
  str_detect(repeat_family_df$gene, "^AluJ") ~ "AluJ",
  str_detect(repeat_family_df$gene, "^AluS") ~ "AluS",
  str_detect(repeat_family_df$gene, "^AluM") ~ "AluM",
  str_detect(repeat_family_df$gene, "^L1")   ~ "L1",
  str_detect(repeat_family_df$gene, "^L2")   ~ "L2",
  str_detect(repeat_family_df$gene, "^L3")   ~ "L3",
  str_detect(repeat_family_df$gene, "^L4")   ~ "L4",
  str_detect(repeat_family_df$gene, "^L5")   ~ "L5",
  str_detect(repeat_family_df$gene, "^CR1")  ~ "CR1",
  str_detect(repeat_family_df$gene, "^HAL1") ~ "HAL1",
  str_detect(repeat_family_df$gene, "_LINE$") ~ "x",
  TRUE ~ "other"
)


# Filter the full data frame to valid genes
repeat_genes <- repeat_family_df$gene
repeat_class_filtered <- repeat_family_df[repeat_family_df$gene %in% repeat_genes, ]

# Split by subfamily
repeat_subfamily_list <- split(repeat_class_filtered$gene, repeat_class_filtered$family)

seurat_obj <- AddModuleScore(seurat_obj, features = repeat_subfamily_list, name = "RepeatSubfamily")

# Rename columns for clarity
subfamily_names <- names(repeat_subfamily_list)
score_cols <- grep("^RepeatSubfamily[0-9]+$", colnames(seurat_obj@meta.data), value = TRUE)
colnames(seurat_obj@meta.data)[match(score_cols, colnames(seurat_obj@meta.data))] <- subfamily_names

repeat_subfam_mat <- t(seurat_obj@meta.data[, subfamily_names])
cor_mat <- cor(t(repeat_subfam_mat), t(tissue_mat), use = "pairwise.complete.obs")
cor_mat = cor_mat[c("AluY", "AluS", "AluJ", "L1", "L2", "L3", "L4", "L5", "HAL1", "CR1"), ]

cor_df <- melt(cor_mat)
colnames(cor_df) <- c("RepeatSubfamily", "Tissue", "Correlation")

# Plot
ggplot(cor_df, aes(x = Tissue, y = RepeatSubfamily, color = Correlation, size = abs(Correlation))) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_classic() +
  labs(title = "Correlation of Repeat Subfamily Scores with Tissue Identity",
       x = "Tissue Type", y = "Repeat Subfamily") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Reshape module scores from wide to long
score_long <- meta %>%
  dplyr::select(cells, gptcelltype_10, all_of(subfamily_names)) %>%
  pivot_longer(
    cols = all_of(subfamily_names),
    names_to = "RepeatSubfamily",
    values_to = "ModuleScore"
  )


ggplot(score_long, aes(x = gptcelltype_10, y = ModuleScore, fill = gptcelltype_10)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~ RepeatSubfamily, scales = "free_y") +
  scale_fill_brewer(palette = "Set3") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "gptcelltype_10", y = "Module Score", title = "Repeat Subfamily Scores by x")



##vln plot
meta <- seurat_obj@meta.data
meta$cells <- rownames(meta)

# Reshape module scores from wide to long
score_long <- meta %>%
  dplyr::select(cells, gptcelltype_10, all_of(subfamily_names)) %>%
  pivot_longer(
    cols = all_of(subfamily_names),
    names_to = "RepeatSubfamily",
    values_to = "ModuleScore"
  )

# Compute median ModuleScore per cluster and subfamily
median_scores <- score_long %>%
  group_by(RepeatSubfamily,gptcelltype_10 ) %>%
  summarise(median_score = median(ModuleScore), .groups = "drop")

# Create a cluster ordering per subfamily
cluster_order <- median_scores %>%
  group_by(RepeatSubfamily) %>%
  arrange(desc(median_score)) %>%
  mutate(order = row_number())

# Merge with main data to add order info
score_long_ordered <- score_long %>%
  left_join(cluster_order, by = c("RepeatSubfamily", "gptcelltype_10"))

# Convert to ordered factor
score_long_ordered$gptcelltype_10 <- factor(
  score_long_ordered$gptcelltype_10,
  levels = unique(score_long_ordered$gptcelltype_10[order(score_long_ordered$order)])
)



##for all
ggplot(score_long_ordered, aes(x = gptcelltype_10, y = ModuleScore, fill = gptcelltype_10)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5, color = "black") +
  facet_wrap(~ RepeatSubfamily, scales = "free_y") +
  scale_fill_brewer(palette = "Set3") +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(x = "Cluster (ordered by median score)", y = "Module Score", title = "Repeat Subfamily Scores by Cluster")


##for each repeat
target_repeat <- "L3"

# Filter and reorder cluster levels by median
score_filtered <- score_long_ordered %>%
  filter(RepeatSubfamily == target_repeat) %>%
  mutate(
    gptcelltype_10 = factor(
      gptcelltype_10,
      levels = score_long_ordered %>%
        filter(RepeatSubfamily == target_repeat) %>%
        group_by(gptcelltype_10) %>%
        summarize(med = median(ModuleScore, na.rm = TRUE)) %>%
        arrange(desc(med)) %>%
        pull(gptcelltype_10)
    )
  )
# Define custom colors
cluster_colors <- c(
  "Fibroblasts" = "yellow", 
  "Hepatic Stellate Cells" = "purple",  # Epithelial Cells
  "Intestinal Epithelial Cells" = "green",  # Endothelial Cells
  "Langerhans Cells" = "orange",  # Myeloid Cells
  "Liver Cells" = "blue",  # Plasma Cells
  "Muscle Cells" = "red" , 
  "Pancreatic Cells" = "pink" , 
  "Plasma Cells" = "brown"  
)

# Then use:
plot = ggplot(score_filtered, aes(x = gptcelltype_10, y = ModuleScore, fill = gptcelltype_10)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5, color = "black") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(
    x = "Cluster (ordered by median score)",
    y = "Module Score",
    title = paste(target_repeat, "Module Scores by Cluster")
  )



pdf("~/Downloads/AAL3.pdf", width = 3, height = 3)  
print(plot)
dev.off()



saveRDS(seurat_obj, file = "./PDAC_moduleScores.rds")
```

#check cluster markers uniqueness for cell type anno
```{r}
seuratObj_cellline = readRDS("0725PDAC_EM_GENEFULL_cellAno.rds")
cluster_marker = fread("PDAC_cluster_markers.txt")
marker_ranked <- cluster_marker %>%
  mutate(rank_score = -log10(p_val_adj) * abs(avg_log2FC)) %>%
  group_by(cluster) %>%
  arrange(desc(rank_score), .by_group = TRUE)
# Get top 10 genes per cluster
top10_genes <- marker_ranked %>%
  slice_head(n = 10)

seuratObj_cellline <- ScaleData(seuratObj_cellline, features = unique(top10_genes$gene))

marker_genes_list <- unique(top10_genes$gene)

DoHeatmap(seuratObj_cellline,
          features = marker_genes_list,
          group.by = "seurat_clusters",
          cells = sample(colnames(seuratObj_cellline), 1000)) +
  theme(axis.text.y = element_text(size = 10))


top10 <- cluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC)

seuratObj_cellline <- ScaleData(seuratObj_cellline, features = unique(top10$gene))

DoHeatmap(seuratObj_cellline,
          features = unique(top10$gene),
          group.by = "seurat_clusters",
          cells = sample(colnames(seuratObj_cellline), 1000)) +
  theme(axis.text.y = element_text(size = 10))

#check specific marker
FeaturePlot(seuratObj_cellline, features = c("ACTA2")) #cluster 0, hepatic stellate cells, no
FeaturePlot(seuratObj_cellline, features = c("HSPH1")) #cluster 1, fibroblast, ok
FeaturePlot(seuratObj_cellline, features = c("APOC3")) #intestinal epithelial cells, cluster 6 and 7
FeaturePlot(seuratObj_cellline, features = c("VWF")) #cluster 5 endothelial cells not plasma cells
FeaturePlot(seuratObj_cellline, features = c("IRF4")) #cluster 5 endothelial cells?
FeaturePlot(seuratObj_cellline, features = c("BIN1")) #cluster 9, muscle cells, ok
FeaturePlot(seuratObj_cellline, features = c("MKI67")) 



```


##umap showing Alu and LINE gene enrichment
```{r}
data_mat <- GetAssayData(seurat_obj, slot = "data")  # normalized log counts
alu_genes <- grep("^Alu", rownames(data_mat), value = TRUE)
alu_mat <- data_mat[alu_genes, ]
alu_bin <- alu_mat > 0
clusters <- Idents(seurat_obj)
alu_gene_count_per_cell <- Matrix::colSums(alu_bin)
df <- data.frame(
  AluGeneCount = alu_gene_count_per_cell,
  Cluster = clusters
)
ggplot(df, aes(x = Cluster, y = AluGeneCount)) +
  geom_violin(trim = FALSE, scale = "width") +
  theme_minimal() +
  labs(title = "Number of Alu Genes Expressed per Cell by Cluster",
       y = "Alu Genes Expressed", x = "Cluster")

#Alu expression
alu_genes <- grep("^Alu", rownames(seurat_obj), value = TRUE)
alu_genes <- intersect(alu_genes, rownames(seurat_obj))
seurat_obj <- AddModuleScore(seurat_obj, features = list(alu_genes), name = "Alu")

plot = FeaturePlot(seurat_obj, features = "Alu1", cols = c("blue", "red"),  raster=FALSE) +
  ggtitle("UMAP - Alu Gene Enrichment")
ggsave("PDACAlu Gene Enrichment.png", plot = plot, bg = "transparent", width = 6, height = 6)

seurat_alu <- subset(seurat_obj, features = alu_genes)

# Identify markers for the cluster
alu_markers <- FindMarkers(seurat_alu, ident.1 = 3, only.pos = TRUE)
alu_markers <- alu_markers[order(alu_markers$avg_log2FC, decreasing = TRUE), ]
head(alu_markers, 10)  # top enriched Alu genes

top_alu <- rownames(alu_markers)[1:6]  # or top 10
DotPlot(seurat_obj, features = top_alu, group.by = "seurat_clusters") + RotatedAxis()


#L1 expression
l1_genes <- grep("^L1", rownames(seurat_obj), value = TRUE)
l1_genes <- intersect(l1_genes, rownames(seurat_obj))  # make sure they exist in object
seurat_obj <- AddModuleScore(seurat_obj, features = list(l1_genes), name = "L1")
FeaturePlot(seurat_obj, features = "L11", cols = c("lightgrey", "red"),  raster=FALSE) +
  ggtitle("UMAP - L1 Gene Enrichment")



 
```

##percentage of Alu overall and per sample
```{r}
seurat_obj = readRDS("PDAC_moduleScores.rds")
repeat_class = fread("/data1/greenbab/users/nij2/ref/all_landscape_rm.csv")
repeat_class = repeat_class[repeat_class$class == "SINE", ]
##whole thing
# Identify repeat and coding genes
repeat_genes <- repeat_class$gene.id
all_genes <- rownames(seurat_obj)
coding_genes <- setdiff(all_genes, repeat_genes)

# Extract raw count matrix (assumes default assay is RNA and counts layer is accessible)
counts <- LayerData(seurat_obj, layer = "counts")  # or use GetAssayData(seurat_obj, slot = "counts") in Seurat v4

# Sum UMI counts across all cells
repeat_umi <- sum(counts[intersect(repeat_genes, rownames(counts)), ])
coding_umi <- sum(counts[intersect(coding_genes, rownames(counts)), ])

# Compute percentages
repeat_fraction <- repeat_umi / (repeat_umi + coding_umi)
coding_fraction <- 1 - repeat_fraction

# Output
cat(sprintf("Repeat fraction (UMIs): %.2f%%\n", repeat_fraction * 100))
cat(sprintf("Coding fraction (UMIs): %.2f%%\n", coding_fraction * 100))



##per sample
library(dplyr)
library(ggplot2)

# Extract counts
counts <- LayerData(seurat_obj, layer = "counts")

# Initialize a dataframe to store results
sample_repeat_fraction <- data.frame()

# Loop over samples
for (sample in unique(seurat_obj$sample_id)) {
  # Cells from this sample
  cells <- colnames(seurat_obj)[seurat_obj$sample_id == sample]
  
  # Subset counts for these cells
  counts_sample <- counts[, cells, drop = FALSE]
  
  # Sum repeat and coding UMIs
  repeat_umi <- sum(counts_sample[intersect(repeat_genes, rownames(counts_sample)), ])
  coding_umi <- sum(counts_sample[intersect(coding_genes, rownames(counts_sample)), ])
  
  # Compute fraction
  repeat_fraction <- repeat_umi / (repeat_umi + coding_umi)
  
  # Store
  sample_repeat_fraction <- rbind(sample_repeat_fraction,
                                  data.frame(sample = sample, repeat_fraction = repeat_fraction))
}


ggplot(sample_repeat_fraction, aes(x = "", y = repeat_fraction)) +
  geom_boxplot(outlier.shape = NA, fill = "lightblue") +
  geom_jitter(width = 0.2, size = 2, color = "darkblue") +
  labs(title = "Alu Fraction per Sample",
       x = "Samples",
       y = "Alu Gene Fraction (nCount)") +
  theme_minimal(base_size = 14)















```


```{r}
library(gtools)
repeat_class_names <- names(repeat_class_list)

# Sum all repeat module scores per cell
seurat_obj@meta.data$Repeat_TotalScore <- rowSums(seurat_obj@meta.data[, repeat_class_names], na.rm = TRUE)
score_df_total <- seurat_obj@meta.data[, c("tissue", "patient", "Repeat_TotalScore")]

# Convert to long format (just one column of scores)
score_long_total <- score_df_total %>%
  rename(ModuleScore = Repeat_TotalScore)

ordered_patients <- mixedsort(unique(score_long_total$patient))
score_long_total$patient <- factor(score_long_total$patient, levels = ordered_patients)

ggplot(score_long_total, aes(x = tissue, y = ModuleScore, fill = tissue)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~ patient, scales = "free_y", nrow = 1) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 9),
    legend.position = "none"
  ) +
  labs(
    x = "Tissue",
    y = "Total Repeat Module Score",
    title = "Total Repeat Expression per Patient by Tissue"
  )


score_df <- seurat_obj@meta.data[, c("tissue", "patient", repeat_class_names)]
score_long <- score_df %>%
  pivot_longer(cols = all_of(repeat_class_names), names_to = "RepeatClass", values_to = "ModuleScore")

# Filter to desired repeat classes
selected_classes <- c("DNA", "LINE", "SINE", "LTR", "rRNA", "Satellite", "Simple_repeat", "sRNA", "tRNA")
score_long_filtered <- score_long %>%
  filter(RepeatClass %in% selected_classes)


# Violin plot per patient, faceted by tissue and colored by repeat class
ggplot(score_long_filtered, aes(x = tissue, y = ModuleScore, fill = tissue)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_grid(RepeatClass ~ patient, scales = "free_y", space = "free_x") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 9),
    legend.position = "none"
  ) +
  labs(x = "Tissue", y = "Module Score", title = "Repeat Class Expression per Patient by Tissue")

```
